Equalizer Programming Guide

### 2008-02-20 Developer/ProgrammingGuide/paper.tex
Programming Guide: add advance chapter on thread synchronization
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

\date{
  \vspace{2cm}
  \begin{figure}[ht]
    \hspace{-4.5cm}\includegraphics[width=22cm]{images/teaser.pdf}
  \end{figure}
  \vspace{1cm}
%  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
  Version 1.1.10 for Equalizer 0.4+svn\\[\medskipamount]
  \today
}

@@

\textcopyright 2007 Eyescale Software GmbH. All rights reserved. No
permission is granted to copy, distribute, or create derivative works
from the contents of this electronic documentation in any manner, in
whole or in part, without the prior written permission of Eyescale
Software GmbH.

@@

\clearpage
\thispagestyle{empty}
\tableofcontents
\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.0.1   & Nov 06, 2007 & Fixed Typo\\
    1.1     & Nov 12, 2007 & Adapted distributed objects\\
    1.1.1   & Nov 16, 2007 & Minor improvements based on lulu.com print\\
    1.1.2   & Nov 19, 2007 & Minor improvements based on feedback\\
    1.1.3   & Nov 23, 2007 & Add OpenGL extension handling\\
    1.1.4   & Nov 26, 2007 & Changes caused by removal of glX event thread\\
    1.1.5   & Dec 10, 2007 & NodeFactory API changes\\
    1.1.6   & Dec 17, 2007 & Add Scalable Rendering chapter\\
    1.1.7   & Dec 20, 2007 & Per-window object manager changes\\
    1.1.8   & Jan 15, 2007 & Detailed window initialization based on
                             latest refactoring\\
    1.1.9   & Jan 16, 2007 & Detail distributed object usage\\
    1.1.10  & Jan 28, 2007 & Add creation thread for NodeFactory entities\\
    \hline
  \end{tabularx}}
\thispagestyle{empty}
\clearpage

@@

@@

\subsection{Execution Model and Thread Safety}

Using threading correctly in OpenGL-based applications is easy with
Equalizer. Equalizer creates one rendering thread for each graphics
card. All task methods for a pipe, and therefore all OpenGL commands,
are executed from this thread. This threading model is the OpenGL
`threading model', which maintains a current context for each
thread. If structured correctly, the application rarely has to take care
of thread synchonization or protection of shared data.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/model.pdf}
  {\caption{\small\label{fModel}Simplified execution model}}
\end{wrapfigure}
The main thread is responsible for maintaining the application logic. It
reacts on user events, updates the data model and requests new frames to
be rendered. It drives the whole application, as shown in \fig{fModel}.

@@

This per-node frame synchronization does not inhibit latency across
rendering nodes. Furthermore, advanced rendering software which
multi-buffers the dynamic parts of the database can disable the per-node
frame synchronization, as explained in \sref{sNodeFrame}. Some scene
graphs do implement multi-buffered data.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \subfigure[]{\label{fSync}
    \includegraphics[width=.25\textwidth]{images/sync}}\hfil
  \subfigure[]{\label{fAsync}
    \includegraphics[width=.25\textwidth]{images/async}}%
  {\caption{\small\label{fSyncAsync}Synchronous and asynchronous execution}}
\end{wrapfigure}
By default, \textsf{Config::finish\-Fra\-me} also synchronizes the
completion of all local rendering tasks for the current frame. This
facilitates porting of exising rendering codes, since the database does
not have to be multi-buffered. Applications such as \textsf{eqPly}, which
do not need this per-node frame synchronization can disable it, as
explained in \sref{sNodeFrame}.

\fig{fSyncAsync} shows the execution of (hypothetical) rendering tasks
without latency (\fig{fSync}) and with a latency of
one frame (\fig{fAsync}). With \textsf{eqPly}, a speedup of 15\% has been
observed on a five-node rendering cluster when using a latency of one frame
instead of no
latency\footnote{\link{http://www.equalizergraphics.com/scalability.html}}. A
latency of one or two frames is normally not perceived by the user.

@@

\subsubsection{Best Practices}

Distributed objects are addressed using session-unique identifiers,
because pointers to other objects can not be distributed directly, since
they have no meaning on remote nodes. The session with which distributed
objects are registered is normally the \textsf{eq::Config}.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/objects.pdf}
  {\caption{\small\label{fObjects}Distributed Data Handling}}
\end{wrapfigure}
Applications distributing their scene graph, rather then relying on a
shared filesystem like \textsf{eqPly}, use the frame data also as an
entry point to their scene graph data structure. \fig{fObjects} shows
one possible implementation, where the identifier and version of the
scene graph root are transported using the frame data. The scene graph
root then serializes and de-serializes his immediate children by
transferring their identifier and current version.

@@

\subsubsection{\label{sNodeFrame}Frame Control}

The application has extended control over the task synchronization
during a frame. Upon \textsf{Config::startFrame}, Equalizer invokes the
\textsf{frameStart} task methods of the various entities. The entities
unlock all its children by calling \textsf{startFrame}, e.g.,
\textsf{Node::frameStart} has to call \textsf{Node::startFrame} in order
to unlock the pipe threads. Note that certain \textsf{startFrame} calls,
e.g., \textsf{Window::startFrame}, are currently empty since the
synchronization is implicit due to the sequential execution within the
thread.

Likewise, \textsf{Config::finishFrame} causes the invokation of the
\textsf{frameFinish} task methods. These task methods unlock their
parents by calling \textsf{releaseFrame}.

The explicit synchronization of child or parent resources allows the
application to optimize the processing, by performing certain, independent
operations when the child or parent resources are already unlocked.

The default thread synchronization synchronizes all
\textsf{Channel::frameDraw} operations on a \textbf{single} node with
the node's main thread. This facilitates porting, since the scene
database does not have to be multi-buffered as all threads of the node
process render the same frame.

The per-node frame synchronization does not break the asynchronous
execution across rendering nodes.  Advanced applications can remove per-node
frame synchronization.

The per-node frame synchronization is achieved through the
\textsf{startFrame\textrightarrow wait\-Fra\-me\-Started} and
\textsf{releaseFrameLocal\textrightarrow waitFrameLocal} synchronization
points. Note that this synchronization is only per-node, different nodes
in the cluster still run asynchronously.

The first synchronization point ensures that the application (node)
thread is done modifiying the data, typically during event
processing. The pipe threads call \textsf{Node::waitFrameStarted} which
blocks until the node calls \textsf{startFrame} in
\textsf{Node::frame\-Start}.

The second pair ensures that after \textsf{Config::finishFrame} all pipe
threads are done rendering the current frame, and therefore do not
access shared data. At this point the application thread can safely
modify scene data. The node's \textsf{frameDrawFinish} waits for all
local pipes to release the synchronization by calling
\textsf{Pipe::release\-Fra\-me\-Lo\-cal}, which happens by default in
\textsf{Pipe::frameDrawFinish}. The \textsf{frameDrawFinish} methods are
called after all \textsf{Channel::frameDraw} of the corresponding thread
have been executed.

\begin{figure}[ht!]\center
  \includegraphics[width=\textwidth]{images/mainloop.pdf}
  {\caption{\small\label{fFrameSync}Synchronization of frame tasks}}
\end{figure}

\fig{fFrameSync} outlines the synchronization for the application, node
and pipe classes. Please note that \textsf{Config::finishFrame} does
block until the current frame has been released locally and until the
frame \textsf{current - latency} has been released by all nodes. The
window and channel synchronization are similar and omitted for
simplicity.

Since \textsf{eqPly} multi-buffers all dynamic data, it can completely
remove frame synchronization by:

\begin{itemize}
\item releasing the local synchronization early in \textsf{Node::frameStart}
\item not waiting for the node to start the frame by not calling
  \textsf{Node::waitFrameStarted} in \textsf{Pipe::frameStart}
\item not waiting for the pipe synchronization in \textsf{Node::frameDrawFinish}
\end{itemize}

Releasing the local synchronization early is done by calling
\textsf{releaseFrameLocal}, and \textsf{Node::frameDrawFinish} is
overwritten with an empty implementation:

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/windowInit.pdf}
  {\caption{\small\label{fWindowInit}Window Initialization}}
\end{wrapfigure}
The initialization sequence uses multiple, overrideable task
methods. The main task method \textsf{configInit} executes a `child'
task method to create the drawable and context. The child task method
depends on the window system of the pipe. The default implementations of
\textsf{configInitGLX}, \textsf{configInitWGL} or \textsf{configInitAGL}
create an on-screen window using OS-specific methods. If the OpenGL
drawable and context were created successfully, \textsf{configInit} calls
\textsf{configInitGL}, which performs the generic OpenGL
state initialization. The default implementation sets up some typical
OpenGL state, e.g., it enables the depth test.

@@

The window-system specific initialization takes into account various
attributes set in the configuration file. Attributes include the size of
various frame buffer attachments (color, alpha, depth, stencil) as well
as other framebuffer properties, such as quad-buffered stereo,
doublebuffering, fullscreen mode and window decorations. Some of the
attributes, such as stereo, doublebuffer and stencil can be set to
\textsf{eq::AUTO}, in which case Equalizer will test for their availability and 
enable them if possible.

The \textsf{eqPly} window initialization function first calls
\textsf{eq::Window::configInit} to use the generic window setup. If that
was successful, it initializes a state object and an overlay logo:

{\footnotesize\begin{lstlisting}
bool Window::configInit( const uint32_t initID )
{
    if( !eq::Window::configInit( initID ))
        return false;

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/render.pdf}
  {\caption{\small\label{fRender}Main Rendering Loop}}
\end{wrapfigure}
The model data is spatially organized in an 3-dimensional
kD-tree\footnote{See also \link{http://en.wikipedia.org/wiki/Kd-tree}}
for efficient view frustum culling. When the model is loaded by
\textsf{Node::config\-Init}, it is preprocessed into the kD-tree and each
node of the tree gets a database range assigned. The root node
has the range [0, 1], its left child [0, 0.5] and its right child [0.5,
1], and so on for all nodes in the tree. The preprocessed model is saved
in a binary format for accelerating subsequent use.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/eventFilter.pdf}
  {\caption{\small\label{fEventProcessing}Event Processing}}
\end{wrapfigure}
Events are received by an event handler. The event handler finds the
\textsf{eq::Window} associated to the event. It then creates a generic
\textsf{WindowEvent}, which holds important event data in a 
window system-independent format. The original event is attached to the
generic window event.

The event handler then passes the window event to
\textsf{Window::processEvent}, which is responsible for either handling
the event locally, or for translating it into a generic
\textsf{ConfigEvent}. The config events are send to the application
thread using \textsf{Config::sendEvent}.

If the event was processed by \textsf{processEvent}, the function has to
return \textsf{true}. If \textsf{false} is returned, the event will be
passed to a previously installed, window-system-specific event handling
function. The default implementation of \textsf{Window::processEvent}
passes most events on to the application.

@@

Later Equalizer versions will introduce \textsf{Pipe::processEvent} and
\textsf{PipeEvent} to communicate pipe-specific events, e.g., monitor
resolution changes.

@@

@@

The main idea behind direct send is to parallelize the costly
recomposition for database (sort-last) decomposition. With each
additional source channel, the amount of pixel data to be composited
grows linearly. When using the simple approach of compositing all frames
on the destination channel, this channel quickly becomes the bottleneck
in the system. Direct send distributes this workload evenly across all
source channels, and thereby keeps the compositing work per channel
constant.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/directSend.pdf}
  {\caption{\small\label{fDirectSend}Parallel Direct Send Compositing}}
\end{wrapfigure}
In direct send compositing, each rendering channel is also responsible
for the sort-last composition\footnote{depth-based for polygonal data or
  sorted-blend for volume data} of one screen-space tile. He receives
the framebuffer pixels for his tile from all the other channels. The
size of one tile decreases linearly with the number of source channels,
which keeps the total amount of pixel data per channel constant.

@@

The frame synchronization through the frame data allows the input frame
to wait for the pixel data to become ready, which is signalled by the
output frame after readback.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/assembly.pdf}
  {\caption{\small\label{fAssembly}Hierarchy of assembly classes}}
\end{wrapfigure}
Readback and assemble operations on frames and images are designed to be
asynchronous. They have a start and finish method for both readback and
assemble to allow the initiation and synchronization of the operation.
Currently, only synchronous readback and assembly using
\textsf{glReadPixels} and \textsf{glDrawPixels} is implemented in the
respective start method of the image. Later versions of Equalizer will
implement asynchronous pixel transfers.

>>>>

\date{
  \vspace{2cm}
  \begin{figure}[ht]
    \hspace{-4.5cm}\includegraphics[width=22cm]{images/teaser.pdf}
  \end{figure}
  \vspace{1cm}
%  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
  Version 1.1.11 for Equalizer 0.4+svn\\[\medskipamount]
  \today
}

@@

\textcopyright 2007-2008 Eyescale Software GmbH. All rights reserved. No
permission is granted to copy, distribute, or create derivative works
from the contents of this electronic documentation in any manner, in
whole or in part, without the prior written permission of Eyescale
Software GmbH.

@@

\clearpage
\thispagestyle{empty}
\tableofcontents
\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.0.1   & Nov 06, 2007 & Fixed Typo\\
    1.1     & Nov 12, 2007 & Adapted distributed objects\\
    1.1.1   & Nov 16, 2007 & Minor improvements based on lulu.com print\\
    1.1.2   & Nov 19, 2007 & Minor improvements based on feedback\\
    1.1.3   & Nov 23, 2007 & Add OpenGL extension handling\\
    1.1.4   & Nov 26, 2007 & Changes caused by removal of glX event thread\\
    1.1.5   & Dec 10, 2007 & NodeFactory API changes\\
    1.1.6   & Dec 17, 2007 & Add Scalable Rendering chapter\\
    1.1.7   & Dec 20, 2007 & Per-window object manager changes\\
    1.1.8   & Jan 15, 2008 & Detailed window initialization based on
                             latest refactoring\\
    1.1.9   & Jan 16, 2008 & Detail distributed object usage\\
    1.1.10  & Jan 28, 2008 & Add creation thread for NodeFactory entities\\
    1.1.11  & Feb 20, 2008 & Add advanced chapter for thread synchronization\\
    \hline
  \end{tabularx}}
\thispagestyle{empty}
\clearpage

@@

\subsubsection{Parallel Rendering}

\begin{wrapfigure}{r}{.618\textwidth}
  \vspace{-1em}
  \includegraphics[width=.618\textwidth]{images/executionFlow.pdf}
  {\caption{\small\label{fExecutionFlow}Parallel Rendering}}
  \vspace{-1em}
\end{wrapfigure}

\fig{fExecutionFlow} illustrates the basic principle of any parallel
rendering application. The typical OpenGL application, for example
GLUT-based applications, has an event loop which redraws the scene,
updates data based on received events, and eventually redraws a new
frame.

Any parallel rendering application keeps this basic execution model and
extends it by separating the rendering code from the main event
loop. This rendering coded is then executed in parallel on different
resources. This model is followed by Equalizer, thus making application
development as easy as possible.

@@

\subsection{\label{sExecution}Execution Model and Thread Safety}

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/model.pdf}
  {\caption{\small\label{fModel}Simplified execution model}}
\end{wrapfigure}
Using threading correctly in OpenGL-based applications is easy with
Equalizer. Equalizer creates one rendering thread for each graphics
card. All task methods for a pipe, and therefore all OpenGL commands,
are executed from this thread. This threading model is the OpenGL
`threading model', which maintains a current context for each
thread. If structured correctly, the application rarely has to take care
of thread synchonization or protection of shared data.

The main thread is responsible for maintaining the application logic. It
reacts on user events, updates the data model and requests new frames to
be rendered. It drives the whole application, as shown in \fig{fModel}.

@@

This per-node frame synchronization does not inhibit latency across
rendering nodes. Furthermore, advanced rendering software which
multi-buffers the dynamic parts of the database can disable the per-node
frame synchronization, as explained in \sref{sNodeFrame}. Some scene
graphs for example do implement multi-buffered data.

@@

By default, \textsf{Config::finish\-Fra\-me} also synchronizes the
completion of all local rendering tasks for the current frame. This
facilitates porting of exising rendering codes, since the database does
not have to be multi-buffered. Applications such as \textsf{eqPly}, which
do not need this per-node frame synchronization can disable it, as
explained in \sref{sNodeFrame}.

\begin{wrapfigure}{r}{.618\textwidth}
  \subfigure[]{\label{fSync}
    \includegraphics[width=.25\textwidth]{images/sync}}\hfil
  \subfigure[]{\label{fAsync}
    \includegraphics[width=.25\textwidth]{images/async}}%
  {\caption{\small\label{fSyncAsync}Synchronous and asynchronous execution}}
\end{wrapfigure}
\fig{fSyncAsync} shows the execution of (hypothetical) rendering tasks
without latency (\fig{fSync}) and with a latency of
one frame (\fig{fAsync}). With \textsf{eqPly}, a speedup of 15\% has been
observed on a five-node rendering cluster when using a latency of one frame
instead of no
latency\footnote{\link{http://www.equalizergraphics.com/scalability.html}}. A
latency of one or two frames is normally not perceived by the user.

@@

\subsubsection{Object Usage for Rendering}

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/objects.pdf}
  {\caption{\small\label{fObjects}Distributed Data Handling}}
\end{wrapfigure}
Distributed objects are addressed using session-unique identifiers,
because pointers to other objects can not be distributed directly, since
they have no meaning on remote nodes. The session with which distributed
objects are registered is normally the \textsf{eq::Config}.

@@

Applications distributing their scene graph, rather then relying on a
shared filesystem like \textsf{eqPly}, use the frame data also as an
entry point to their scene graph data structure. \fig{fObjects} shows
one possible implementation, where the identifier and version of the
scene graph root are transported using the frame data. The scene graph
root then serializes and de-serializes his immediate children by
transferring their identifier and current version.

@@

The two remaining functions in the node relax the thread synchronization
between the node and pipe threads, since all dynamic data is
multi-buffered in \textsf{eqPly}. \sref{sThreads} provides a detailed
explanation of thread synchronization in Equalizer:

@@

The initialization sequence uses multiple, overrideable task
methods. The main task method \textsf{configInit} executes a `child'
task method to create the drawable and context. The child task method
depends on the window system of the pipe. The default implementations of
\textsf{configInitGLX}, \textsf{configInitWGL} or \textsf{configInitAGL}
create an on-screen window using OS-specific methods. If the OpenGL
drawable and context were created successfully, \textsf{configInit} calls
\textsf{configInitGL}, which performs the generic OpenGL
state initialization. The default implementation sets up some typical
OpenGL state, e.g., it enables the depth test.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/windowInit.pdf}
  {\caption{\small\label{fWindowInit}Window Initialization}}
\end{wrapfigure}
The window-system specific initialization takes into account various
attributes set in the configuration file. Attributes include the size of
the various frame buffer planes  (color, alpha, depth, stencil) as well
as other framebuffer attributes, such as quad-buffered stereo,
doublebuffering, fullscreen mode and window decorations. Some of the
attributes, such as stereo, doublebuffer and stencil can be set to
\textsf{eq::AUTO}, in which case Equalizer will test for their availability and 
enable them if possible.

For the window-system specific initialization, \textsf{eqPly} uses the
default Equalizer implemenation. The \textsf{eqPly} window
initialization only overrides the OpenGL-specific initialization
function in order to initialize a state object and an overlay logo. This
function is only called if an OpenGL context was created and made current:

{\footnotesize\begin{lstlisting}
bool Window::configInitGL( const uint32_t initID )
{
    if( !eq::Window::configInitGL( initID ))
        return false;

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \vspace{-2em}
  \includegraphics[width=.618\textwidth]{images/render.pdf}
  {\caption{\small\label{fRender}Main Rendering Loop}}
  \vspace{-2em}
\end{wrapfigure}
The model data is spatially organized in an 3-dimensional
kD-tree\footnote{See also \link{http://en.wikipedia.org/wiki/Kd-tree}}
for efficient view frustum culling. When the model is loaded by
\textsf{Node::config\-Init}, it is preprocessed into the kD-tree and each
node of the tree gets a database range assigned. The root node
has the range [0, 1], its left child [0, 0.5] and its right child [0.5,
1], and so on for all nodes in the tree. The preprocessed model is saved
in a binary format for accelerating subsequent use.

@@

Events are received by an event handler. The event handler finds the
\textsf{eq::Window} associated to the event. It then creates a generic
\textsf{WindowEvent}, which holds important event data in a 
window system-independent format. The original event is attached to the
generic window event.

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/eventFilter.pdf}
  {\caption{\small\label{fEventProcessing}Event Processing}}
\end{wrapfigure}
The event handler then passes the window event to
\textsf{Window::processEvent}, which is responsible for either handling
the event locally, or for translating it into a generic
\textsf{ConfigEvent}. The config events are send to the application
thread using \textsf{Config::sendEvent}.

If the event was processed by \textsf{processEvent}, the function has to
return \textsf{true}. If \textsf{false} is returned, the event will be
passed to a previously installed, window-system-specific event handling
function. The default implementation of \textsf{Window::pro\-cess\-Event}
passes most events on to the application.

@@

Later Equalizer versions will introduce \textsf{Pipe::pro\-cess\-Event} and
\textsf{PipeEvent} to communicate pipe-specific events, e.g., monitor
resolution changes.

@@

\subsection{\label{sThreads}Multi-Threading and Synchronization}

Equalizer applications use multiple, asynchronous execution threads. The
default execution model is focused on making the porting of existing
applications as easy as possible, as described in \sref{sExecution}. The
default, per-node thread synchronization provided by Equalizer can
gradually be relaxed by advanced applications to gain better performance
through higher asynchronicity.

This section explains the multi-threading in Equalizer in detail and
gives advice on how to optimize applications for performance.

\subsubsection{Threads}

The application or node main thread is the primary thread of each
process and executes the \textsf{main} function. The application and
render clients initialize the local node for communications with other
nodes, including the server, using \textsf{Client::initLocal}.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/threads.pdf}
  {\caption{\small\label{fThreads}Threads within one node process}}
\end{wrapfigure}
During this initialization, Equalizer creates and manages two threads
for communication, the receiver thread and the command thread. Normally
no application code is executed from these two threads.

The receiver thread manages the network connections to other nodes and
receives data. It dispatches the received data either to the application
threads, or to the command thread.

The command thread processes Equalizer-related request from other nodes,
for example during \textsf{eqNet::Object} mapping. In some special cases
the command thread executes application code, for example when a remote
node maps a static or unbuffered object,
\textsf{Object::getInstanceData} is called from the command thread.

The receiver and command thread are terminated when the application
stops network communications using \textsf{Client::exitLocal}.

During config initialization, one pipe thread is created for each
pipe. The pipe threads execute all render task methods for this pipe,
and therefore executes the application's rendering code. The pipe
threads are terminated during \textsf{Config::exit}.

The rest of this section discusses the thread synchronization between
the main thread and the pipe threads.

\subsubsection{\label{sNodeFrame}Thread Synchronization Model}

The node main thread and the pipe threads are synchronized with each
other, so that all draw task methods of one rendered frame are executed
at the same time. This model allows to use the same database for
rendering, and safe modifications of this database are possible from the
node thread, since the pipe threads do not execute any rendering tasks
between frames.

Please note that the default thread synchronization synchronizes all
\textsf{Channel::frameDraw} operations on a \textbf{single} node with
the node's main thread. The per-node frame synchronization does not
break the asynchronous execution across nodes. Advanced applications can
remove the per-node frame synchronization.

The application has extended control over the task synchronization
during a frame. Upon \textsf{Config::startFrame}, Equalizer invokes the
\textsf{frameStart} task methods of the various entities. The entities
unlock all their children by calling \textsf{startFrame}, e.g.,
\textsf{Node::frameStart} has to call \textsf{Node::startFrame} in order
to unlock the pipe threads. Note that certain \textsf{startFrame} calls,
e.g., \textsf{Window::startFrame}, are currently empty since the
synchronization is implicit due to the sequential execution within the
thread.

Each entity uses \textsf{waitFrameStarted} to block on the parent's
\textsf{startFrame}, e.g., \textsf{Pipe::\-frame\-Start} calls
\textsf{Node::waitFrameStarted} to wait for the corresponding
\textsf{Node::start\-Fra\-me}. This explicit synchronization allows to
update non-critical data before synchronizing with
\textsf{waitFrameStarted}, or after unlocking using
\textsf{start\-Fra\-me}. \fig{fFrameSync} illustrates this
synchronization model.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/frameSync.pdf}
  {\caption{\small\label{fFrameSync}Per-Node Frame Synchronization}}
\end{wrapfigure}
At the end of the frame, two similar sets of synchronization methods are
used. The first set synchronizes the local execution, while the second
set synchronizes the global execution.

The local synchronization consists of \textsf{releaseFrameLocal} to
unlock the local frame, and of \textsf{waitFrameLocal} to wait for the
unlock. For the default synchronization model, Equalizer uses the task
method \textsf{frameDrawFinish} which is called on each resource entity
after the last \textsf{Channel::frameDraw} invocation for this
frame. Consequently, \textsf{Pipe::frameDrawFinish} calls
\textsf{Pipe::releaseFrame\-Lo\-cal} to signal that it is done drawing the
current frame, and \textsf{Node::frameDrawFinish} calls
\textsf{Pipe::waitFrameLocal} for each of its pipes to block the node
thread until the current frame has been drawn.

\fig{fFrameSync} illustrates the local frame synchronization. By
removing the calls to \textsf{waitFrameLocal}, the node thread can
disable the synchronization with the pipe threads, which causes the node
operations to overlap with the rendering.

The second, global synchronization is used for the frame completion
during \textsf{Config::finishFrame}, which causes \textsf{frameFinish}
to be called on all entities, passing the oldest frame number, i.e., frame
\textsf{current-latency}. The \textsf{frameFinish} task methods have to
call \textsf{releaseFrame} to signal that the entitity is done with the
frame. The release causes the parent's \textsf{frameFinish} to be
invoked, which is so far synchronized internally. Once all
\textsf{Node::releaseFrame} have been called,
\textsf{Config::finishFrame} returns.

\begin{figure}[ht!]\center
  \includegraphics[width=\textwidth]{images/mainloop.pdf}
  {\caption{\small\label{fMainLoop}Synchronization of frame tasks}}
\end{figure}

\fig{fMainLoop} outlines the synchronization for the application, node
and pipe classes for an application node and one render client. Please
note that \textsf{Config::finishFrame} does block until the current
frame has been released locally and until the frame \textsf{current -
  latency} has been released by all nodes. The window and channel
synchronization are similar and omitted for simplicity.

Since \textsf{eqPly} multi-buffers all dynamic data, it completely
removes frame synchronization by:

\begin{itemize}
\item releasing the local synchronization early in \textsf{Node::frameStart}
\item not waiting for the node to start the frame by not calling
  \textsf{Node::waitFrame\-Star\-ted} in \textsf{Pipe::frameStart}
\item not waiting for the pipe synchronization in
  \textsf{Node::frameDrawFinish} by not calling \textsf{Pipe::waitFrameLocal}
\end{itemize}

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/directSend.pdf}
  {\caption{\small\label{fDirectSend}Parallel Direct Send Compositing}}
\end{wrapfigure}
The main idea behind direct send is to parallelize the costly
recomposition for database (sort-last) decomposition. With each
additional source channel, the amount of pixel data to be composited
grows linearly. When using the simple approach of compositing all frames
on the destination channel, this channel quickly becomes the bottleneck
in the system. Direct send distributes this workload evenly across all
source channels, and thereby keeps the compositing work per channel
constant.

In direct send compositing, each rendering channel is also responsible
for the sort-last composition\footnote{depth-based for polygonal data or
  sorted-blend for volume data} of one screen-space tile. He receives
the framebuffer pixels for his tile from all the other channels. The
size of one tile decreases linearly with the number of source channels,
which keeps the total amount of pixel data per channel constant.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/assembly.pdf}
  {\caption{\small\label{fAssembly}Hierarchy of assembly classes}}
\end{wrapfigure}
The frame synchronization through the frame data allows the input frame
to wait for the pixel data to become ready, which is signalled by the
output frame after readback.

@@

Readback and assemble operations on frames and images are designed to be
asynchronous. They have a start and finish method for both readback and
assemble to allow the initiation and synchronization of the operation.
Currently, only synchronous readback and assembly using
\textsf{glReadPixels} and \textsf{glDrawPixels} is implemented in the
respective start method of the image. Later versions of Equalizer will
implement asynchronous pixel transfers.
